---
title: "Hydrolytic Extracellular Enzyme Pre-test Calculations"
author: "Hava Blair"
date: "June 18, 2020"
output: html_document
params: 
  round: 3
  lab_date: 20200707
---
## Enzyme Pre-test Calculations 

```{r, packages, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr:: opts_chunk$set(tidy = TRUE)
knitr:: opts_chunk$set(tidy.opts = list(width.cutoff=55))

## Load Packages 
library(tidyverse)
library(plater)
library(glue)
library(broom)
library(hms)
library(ggrepel)

```

## Read in plate templates

You need one completed template file for each plate, in CSV format. See https://github.com/havablair/plater-prep 

  
## File organization 

Place template files for each round in their own subfolder named "plater-templates-round#", filling in # with the appropriate numeral.

Now you are ready to read in your files. If you get an error, check whether you have any extra CSV files in the designated folder.

## Read in plate templates & metadata 

```{r files}
file.names <- dir(glue("./data/temp/plater-templates-round{params$round}/"))
file.paths<- paste0(glue("./data/temp/plater-templates-round{params$round}/"), file.names)
plates <- read_plates(file.paths)
colnames(plates) <- c("plate", "wells", "id", "bad_wells", "conc_uM", "fluor")

plates

# To check that all plates read in correctly
head(plates)
unique(plates$plate) 

#read in plate metadata
plate_metadata <- read.csv(glue("./data/metadata/eea_plate_metadata_pretest{params$round}.csv"), stringsAsFactors = FALSE)

colnames(plate_metadata) <- c("plate", "sample_id", "ph_buffer", "moist_soil_mass_g", "soil_water_content", "vol_buffer_ml", "substrates", "time_soil_added", "time_naoh_added", "time_plate_read")

head(plate_metadata)
```

# Metadata calculations

## Incubation time, soil oven dry equivalent
```{r}

# parse the time columns so we can do math on them
plate_metadata <- plate_metadata %>% 
  mutate(time_soil_added = parse_hm(time_soil_added), 
         time_naoh_added = parse_hm(time_naoh_added), 
         time_plate_read = parse_hm(time_plate_read))

# calculate incubation time and convert from seconds to hours
plate_metadata <- plate_metadata %>% 
  mutate(inc_time_hr = as.numeric(((time_plate_read - time_soil_added)/60/60)))


#calculate soil oven dry equivalent 

plate_metadata <- plate_metadata %>% 
  mutate(soil_ov_dry_eq_g = moist_soil_mass_g - (moist_soil_mass_g * soil_water_content))

plate_metadata

```

# Initial Quality Control 

## Remove empty wells and known bad wells


```{r badwells}
#Remove empty wells
no_missing <- subset(plates, !is.na(id) & id != 0 & id != ".")

# Remove bad wells and keep good ones
no_bad <- subset(no_missing, is.na(bad_wells))

no_missing$bad_wells

#Create a dataframe with the details about which wells were removed as "bad".  Can write this to CSV now, or wait until after later QC checks to compile a list of samples that need to be redone.
bad_list <- subset(no_missing, !is.na(bad_wells))

wells_removed <- nrow(plates) - nrow(no_bad)

glue("Removed {wells_removed} wells that were missing or bad")

clean_data <- no_bad

clean_nested <- clean_data %>% 
  group_by(plate) %>% 
  nest()

head(clean_nested)

```

# Calculate plate blank and homogenate blank (hombl)

```{r}
# function to calculate plate blank mean fluor
get_plate_blank <- function(data){
  buf <- data %>% 
    filter(id == "Buf")
    mean(buf$fluor)
}

# apply function to nested dataframe
plate_blanks <- clean_nested %>% 
  mutate(plate_blank = map_dbl(data, get_plate_blank))

# function to calculate homogenate blank (Buf +s) mean fluor 
get_hombl <- function(data){
  buf_soil <- data %>% 
    filter(id == "Buf+s")
    mean(buf_soil$fluor)
}

# apply function to nested dataframe
hombl_calcs <- plate_blanks %>% 
  mutate(hombl = map_dbl(data, get_hombl))

hombl_calcs$data

```


## MUB Standard calculations (for emission coeff calc)

Calculate the mean fluorescence for each concentration of MUB standard (no soil).  Subtract the plate blank to calculate the corrected fluorescence value for the MUB standards. 

```{r}
mub_std_fun <- function(data){
  data %>%
    select(wells, id, fluor, conc_uM) %>% 
    filter(str_detect(wells, "10")) %>% 
    group_by(id) %>% 
    summarise(mub_std_mean = mean(fluor), n = n()) #Add CV here?
}

mub_stds <- hombl_calcs %>% 
  mutate(mub_std_values = map(data, mub_std_fun))

#Can always include an "unnest" option here for quick viewing of the plate numbers and standards


#Subtract the plate blank from the fluorescence values for each averaged concentration

corr_stds <- mub_stds %>% 
  unnest(mub_std_values) %>% 
  mutate(corr_fluor = mub_std_mean - plate_blank, conc_uM = as.numeric(str_replace(id, "MUB", ""))) %>% 
  nest(std_data = c(id, mub_std_mean, n, corr_fluor, conc_uM))

```

## MUB + Soil Standards 

Calculate the mean fluorescence of MUB + soil at each MUB concentration and correct the fluor using the homogenate blank values.

```{r}

#Average the MUB+soil wells  to get mean fluorescence for each concentration and calculate corrected value by subtracting the fluor of the homogenate blank.
quench_std_fun <- function(data, hombl){
  data %>%
    select(wells, id, fluor, conc_uM) %>% 
    filter(str_detect(wells, "09")) %>% 
    mutate(conc_uM = as.numeric(str_replace(conc_uM, "mub_", ""))) %>%
    group_by(conc_uM) %>% 
    summarise(quench_fluor = mean(fluor), n = n(), sd = sd(fluor), cv = (sd(fluor)/mean(fluor))*100, corr_fluor = quench_fluor - hombl) 
}

quench_stds <- corr_stds %>% 
  mutate(quench_std_values = map2(data, hombl, quench_std_fun))

```
# Linear models 

##Linear model functions 

```{r lm-fun}

#function to run a linear model x = MUB concentration, y = mean fluorescence

lm_mod_ftn <- function(df){
  lm(corr_fluor ~ conc_uM, data = df)
}

# functions to extract linear model details calculated above into a nicer format for putting in our graphs

b_fun <- function(mod){
  coefficients(mod)[[1]]
}

slope_fun <- function(mod){
  coefficients(mod)[[2]]
}

r_sq_fun <- function(mod){
  summary(mod)[["r.squared"]]
}

 max_fluor_fun <- function(data){
   max(data$corr_fluor)
 }

```


## Linear model calculations 

```{r lm-calcs}

#calculate linear model for both standard curves (MUB standards (emission) and MUB stds +soil (quench))
lm_calcs <- quench_stds %>% 
  mutate(emis_lm = map(std_data, lm_mod_ftn), quench_lm = map(quench_std_values, lm_mod_ftn))

#Extract linear model details from the MUB standard (emission) linear model
emis_lm_details <- lm_calcs %>% 
  mutate(intcpt_emis = map_dbl(emis_lm, b_fun), slope_emis = map_dbl(emis_lm, slope_fun), r_squared_emis = map_dbl(emis_lm, r_sq_fun), max_fluor_emis = map_dbl(std_data, max_fluor_fun))

#nest the linear model details in a dataframe
nest_emis_stats <- emis_lm_details %>% 
  nest(lm_stats_emis = c(intcpt_emis, slope_emis, r_squared_emis, max_fluor_emis))

#Extract linear models details from the MUB std + soil (quench) linear model
quench_lm_details <- nest_emis_stats %>% 
  mutate(intcpt_quench = map_dbl(quench_lm, b_fun), slope_quench = map_dbl(quench_lm, slope_fun), r_squared_quench = map_dbl(quench_lm, r_sq_fun), max_fluor_quench = map_dbl(quench_std_values, max_fluor_fun))

#nest the linear model details in a dataframe
nest_quench_stats <- quench_lm_details %>% 
  nest(lm_stats_quench = c(intcpt_quench, slope_quench, r_squared_quench, max_fluor_quench))

```


## Plot MUB Standard Curve (emission) 

```{r std-curve, fig.height = 4, fig.width=5}
#Function to make standard curve plots

plot_pts_fun <- function(std_data, lm_stats_emis, plate){
  g <- ggplot() + 
  geom_point(data = std_data, aes(x = conc_uM, y = corr_fluor)) + 
    geom_abline(data = lm_stats_emis, aes(slope = slope_emis, intercept = intcpt_emis)) +
    geom_text(data = lm_stats_emis, aes(x = 1.0, y = 8000, label = paste("R^2=", round(r_squared_emis, digits = 4))), inherit.aes = FALSE) +
    geom_text(data = lm_stats_emis, aes(x = 1.0, y = 10000, label = paste(" y= ", round(slope_emis, digits = 2), "x", " + ", round(intcpt_emis, digits = 2))), inherit.aes = FALSE) + 
    labs(title = glue("{plate} buffer control standard curve"))
  return(g)
}


plot_test <- nest_quench_stats %>% 
  mutate(plot_buffer_control = pmap(list(std_data, lm_stats_emis, plate), plot_pts_fun))

plot_test$plot_buffer_control


```

## Plot MUB standard + soil standard curve (quench)

```{r soil-std-curve}
plot_homog_fun <- function(quench_std_values, lm_stats_quench, plate){
  g <- ggplot() + 
  geom_point(data = quench_std_values, aes(x = conc_uM, y = corr_fluor)) + 
    geom_abline(data = lm_stats_quench, aes(slope = slope_quench, intercept = intcpt_quench)) +
    geom_text(data = lm_stats_quench, aes(x = 1.0, y = (max_fluor_quench), label = paste("R^2=", round(r_squared_quench, digits = 3))), inherit.aes = FALSE) +
    geom_text(data = lm_stats_quench, aes(x = 1.0, y = (max_fluor_quench - 500), label = paste(" y= ", round(slope_quench, digits = 2), "x", " + ", round(intcpt_quench, digits = 2))), inherit.aes = FALSE) + 
    labs(title = glue("{plate} homogenate control standard curve"))
  return(g)
}

homog_plots <- plot_test %>% 
  mutate(plot_homog_control = pmap(list(quench_std_values, lm_stats_quench, plate), plot_homog_fun))

homog_plots$plot_homog_control

```

# Coefficient calculations 

## Calculate emission coefficient

The emission coefficient is the slope (m) from plate b (fluorescence vs MUB conc NO SOIL) divided by the standard well volume - see German et al. 2011 corrigendum. 

German, D. P., Weintraub, M. N., Grandy, A. S., Lauber, C. L., Rinkes, Z. L., & Allison, S. D. (2012). Corrigendum to “Optimization of hydrolytic and oxidative enzyme methods for ecosystem studies” [Soil Biol. Biochem. 43 (2011) 1387–1397]. Soil Biology and Biochemistry, 44(1), 151. https://doi.org/https://doi.org/10.1016/j.soilbio.2011.11.002


Pay attention to units! 
**slope (m) units** from the standard curves = fluor/uM = fluor/ (umol/L) = fluor L / umol = fluor mL / nmol

**assay volume units** are 250uL = 0.00025 L = 0.250 mL 

**emission coefficient** units are fluorescence/nmol. 

```{r, emission-coefficient}

# emission coefficient units are fluorescence / nmol
stdwell_vol_ml <- 0.250

emis_calc <- homog_plots %>% 
  mutate(emis_coeff = map_dbl(lm_stats_emis, ~ as.numeric(.x[["slope_emis"]])/stdwell_vol_ml))

```

### Calculate the quench coefficient 

The quench coefficient is calculated as the slope of the quench standard curve divided by the slope of the buffer control (emission) standard curve.

**Add more information about what the quench coefficient actually is...** 

```{r quench-coefficient}

quench_coeff <- emis_calc %>% 
  mutate(quench_coeff = map2_dbl(lm_stats_quench, lm_stats_emis, ~ as.numeric(.x[["slope_quench"]])/as.numeric(.y[["slope_emis"]])))

```



## Calculate means for assay wells

```{r}
#select data: plate blank, homogenate blank (hombl), emission coefficient, and quench coefficient
plate_data <- quench_coeff %>% 
  select(plate, plate_blank, hombl, emis_coeff, quench_coeff)

plate_data_df <- as.data.frame(plate_data)

#Calculate the mean fluorescence for the assay wells and substrate control wells
mean_fluor_calcs <- clean_data %>% 
  filter(str_detect(id, "Assay") | str_detect(id, "blank")) %>% 
  group_by(plate, id, conc_uM) %>% 
  summarise(fluor_mean = mean(fluor), n = n(), sd = sd(fluor, na.rm = T), cv = sd(fluor, na.rm = T)/mean(fluor)*100)

#QC check - plot dist of each sample and enzyme to see what's normal?

# format dataframe "wide" so that substrate controls are in columns adjacent to assay wells for each enzyme

wide <- mean_fluor_calcs %>% 
  pivot_wider(id_cols = c(plate, id, conc_uM), names_from = id, values_from = fluor_mean, names_prefix = "mean_fluor_", names_sep = "_")

# join coeffs to assay/substrate well calculations 
wide_coeffs <- left_join(wide, plate_data_df, by = "plate")

``` 

##Calculate net fluorescence 

Net fluorescence = ((assay-homogenate control) / quench coefficient) - (substrate control-plate blank)

```{r, net-fluor-calc}

net_fluor_calcs <- wide_coeffs %>% 
  mutate(BG_net_fluor = (((`mean_fluor_B-Assay` - hombl)/quench_coeff)-(`mean_fluor_B-sub-blank` - plate_blank)), 
         Cello_net_fluor = (((`mean_fluor_C-Assay` - hombl)/quench_coeff)-(`mean_fluor_C-sub-blank` - plate_blank)),  
         NAG_net_fluor = (((`mean_fluor_N-Assay` - hombl)/quench_coeff)-(`mean_fluor_N-sub-blank` - plate_blank)),
         P_net_fluor = (((`mean_fluor_P-Assay` - hombl)/quench_coeff)-(`mean_fluor_P-sub-blank` - plate_blank)))

```


##Calculate activity 

Activity(nmol/g^-1 h^-1) = [Net fluor x buffer vol (mL)] / [emis coeff x homogenate vol(mL) x time(h) x soil mass(g)]

Activity units are nmol / g^-1 hr^-1

```{r, activity-calc}

#buffer vol is the volume of buffer used to make the soil slurries. In the pre-test, this is 25mL.  In the regular protocol, it is 50mL
buffer_vol_ml <- 25

#vol of homogenate added to the assay wells
homogenate_vol_ml <- 0.200 

#bring in plate metadata
add_metadata <- left_join(net_fluor_calcs, plate_metadata, by = "plate")

activity_calcs <- add_metadata %>% 
  mutate(BG_activity = (BG_net_fluor * buffer_vol_ml) / (emis_coeff * homogenate_vol_ml * inc_time_hr * soil_ov_dry_eq_g), 
         Cello_activity = (Cello_net_fluor * buffer_vol_ml) / (emis_coeff * homogenate_vol_ml * inc_time_hr * soil_ov_dry_eq_g),
         NAG_activity = (NAG_net_fluor * buffer_vol_ml) / (emis_coeff * homogenate_vol_ml * inc_time_hr * soil_ov_dry_eq_g), 
          P_activity = (P_net_fluor * buffer_vol_ml) / (emis_coeff * homogenate_vol_ml * inc_time_hr * soil_ov_dry_eq_g))

activity_long <- activity_calcs %>% 
  select(plate, conc_uM, BG_activity, Cello_activity, NAG_activity, P_activity, ph_buffer, sample_id) %>% 
  pivot_longer(cols = c(BG_activity, Cello_activity, NAG_activity, P_activity), names_to = "enzyme", values_to = "activity") %>% 
  mutate(enzyme = str_replace(enzyme, "_activity", ""))


#get rid of rows with activity = NA (produced because each plate only measures 2 of 4 enzymes, but the formula above runs the calculation regardless)
activity_long <- activity_long %>% 
  filter(!is.na(activity)) %>% 
  mutate(conc_uM = as.numeric(str_replace(conc_uM, "sub_", "")))

activity_long %>% 
  filter(activity < 0 )

# 1 activity is negative (<0).  It is on plate 3_NP_ST2_7, conc = 4.8uM, N enzyme, activity = -0.265.  Consider flagging and setting to 0? 

write.csv(activity_long, glue("./results/calcs/pretest_eea_{params$round}_{params$lab_date}_calcs.csv"), row.names = F)

```

