---
title: 'Pretest Results: CIG Hydrolytic Extracelleular Enzyme Assays'
author: "Hava Blair"
date: "July 1, 2020"
output:
  pdf_document:
    df_print: kable
  html_document: default  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(glue)
library(ggrepel)
```

# About the pre-test 2 samples

SW1_SH_A_3 = MN River Valley, pH 6.5  
RRV2_SH_A_3 = Red River Valley, pH 8  

I performed the pre-test on these samples using buffers at the soil pH for that sample (6.5 or 8).  I also tested both samples with a neutral (7) buffer.  

# Pre-test data 

Pre-test data was processed with the R-Markdown document "eea_pretest_calcs.Rmd"
```{r data,}
pre_data <- read.csv("./results/calcs/pretest_eea_3_20200707_calcs.csv", stringsAsFactors = FALSE)

# set ph_buffer col as factor 
pre_data$ph_buffer <- as.factor(pre_data$ph_buffer)

head(pre_data)
```
\newpage

# Phosphatase

```{r}

# P Enzyme 

pre_data %>% 
  filter(enzyme == "P") %>% 
  ggplot() +
  geom_point(aes(x = conc_uM, y = activity, color = ph_buffer)) + 
  theme(legend.position = "bottom") + 
  labs(title = "Phosphatase activity as a function of substrate concentration & buffer pH") + 
    facet_wrap(vars(sample_id))
``` 

## Red River Valley
At pH 7, we appear to have saturating substrate concentrations starting around ~ 75 uM, similar to pre-test 1. At substrate concentrations of 300uM, activity for pH 8 is ~90, while at pH 7 it is ~40 (~45% difference)

## MN River Valley
Activities are similar for both pH 6.5 and pH 7.  Appear to reach saturation at >210uM substrate concentration. 

\newpage
# N-acetylglucosaminidase (NAGase)
```{r}
# NAG Enzyme 

pre_data %>% 
  filter(enzyme == "NAG") %>% 
  ggplot() +
  geom_point(aes(x = conc_uM, y = activity, color = ph_buffer)) + 
  theme(legend.position = "bottom") + 
  labs(title = "N-acetylglucosaminidase (NAGase) activity as a function of substrate concentration & buffer pH") + 
    facet_wrap(vars(sample_id))

``` 

## Red River Valley 

Substantially lower activity in the RR sample compared to the SW sample.  But activities at pH 7 vs pH 8 for this sample are pretty similar

## MN River Valley

Activities at pH 6.5 is substantially higher at substrate concentrations 150uM and 210uM.  However at 300uM activites are simlar for pH 6.5 and 7. 

\newpage
# Cellobiohydrolase

```{r}
# Cello Enzyme 

pre_data %>% 
  filter(enzyme == "Cello") %>% 
  ggplot() +
  geom_point(aes(x = conc_uM, y = activity, color = ph_buffer)) + 
  theme(legend.position = "bottom") + 
  labs(title = "Cellobiohydrolase activity as a function of substrate concentration & buffer pH") + 
    facet_wrap(vars(sample_id))
``` 

## Red River Valley 

Somewhat perplexing curve shape here - looks like activities peek around 75-100uM and then decline >150uM for both pHs?   

## MN River Valley 

Activity consistently higher at pH 6.5 compared to pH 7 except at the lowest substrate concentrations.  At 300uM, our typical "saturating" condition for these assays, activity at pH 6.5 is ~13, while at pH 7 it is  ~7.


\newpage
# B-glucosidase 
```{r}
# BG Enzyme 

pre_data %>% 
  filter(enzyme == "BG") %>% 
  ggplot() +
  geom_point(aes(x = conc_uM, y = activity, color = ph_buffer)) + 
  theme(legend.position = "bottom") + 
  labs(title = "B-glucosidase activity as a function of substrate concentration & buffer pH") + 
    facet_wrap(vars(sample_id))


  
```

## Red River Valley 
Activities consistently higher at pH 7 than at pH 8.  At 300uM activity at pH 7 is ~75, while at pH 8 it is 45.  

## SW (MN River Valley) 
Activity consistently higher at pH 6.5 than at pH 7.  

At pH 7, we appear to reach saturating substrate concentrations at >150uM.  

# Overall 
For NAG, the change in buffer pH does not appear to have a big impact on actvity measured at substrate concentrations of 300uM for either RR or SW samples.  

Phosphatase activities for sample RR2 are 45% higher at pH 8 compared to pH 7.  In contrast, for sample SW1 not much difference between pH 6.5 and pH 7. 

BG activity - pH makes a big difference (50%+) in activity for both samples 

Cello activity - at substrate concentrations of 300uM, not much difference in activity for the RR2 sample between pH 7 and 8, but the curve shape overall looks a bit weird. For the SW sample, activity is ~50% at pH 6.5.  
